Explaining probabilistic predictions on the simplex with Shapley compositions

### 2024-01-03 paper/main.tex
edit paper

Let $f:\mathcal{X}\to\mathbb{R}$ be a learned model one want to \emph{locally} explain where $f(\bm{x})$ is the prediction on the instance $\bm{x}\in\mathcal{X}$. Let $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data\footnote{In practice, this is usually unknown but the expectation will be replaced by empirical samplings.}. Let $S\subseteq \mathcal{I}=\{1,2,\dots d\}$, where $d$ is the number of features that composes an instance $\bm{x}\in\mathcal{X}$, be a subset of indices. $\bm{x}_S$ refers to an instance $\bm{x}$ restricted to the features indicated by the indices in $S$.

When an instance $\bm{x}$ is observed, the expected value of the prediction is simply $\mathbb{E}[f(\bm{x}) \mid \bm{x}] = f(\bm{x})$. However, when only $\bm{x}_S$ is given, i.e. part of the features, there is uncertainty about the other features and we therefore compute the expected prediction given $\bm{x}_S$: $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$. The contribution of the feature indexed by $i \notin S$ in the prediction $f(\bm{x})$ given the known features indexed by $S$ is given by:
\begin{equation}
  \label{eq:contrib}
  c_{f,\bm{x},\text{Pr}}(i,\bm{X}_S) = v_{f,\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) - v_{f,\bm{x},\text{Pr}}(\bm{X}_S),
\end{equation}
where $v$ is known as the value function:
\begin{equation}
  \label{eq:valuefunction}
  \begin{aligned}
    v_{f,\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathbb{R},\\
    S &\mapsto \mathbb{E}_\text{Pr}[f(\bm{x})\mid \bm{x}_S],
  \end{aligned}
\end{equation}
where $2^{\mathcal{X}}$ is the set of all subsets of $I$. This measure the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$. The whole contribution of the $i$th feature is computing by averaging this quantity over all possible coalitions as follow:
\begin{equation}
  \phi_{f,\bm{x},\text{Pr}}(i) = \frac{1}{d!} \sum_{\pi}c_{f,\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}),
\end{equation}
where $\pi$ is a permutation of the set $I$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$. For better clarity, the subscript $_{f,\bm{x},\text{Pr}}$ wil be dropt from the equations.

This quantity is known as the Shapley value for the $i$th feature. It comes from cooperative game theory and is known to be only quantity respecting a set of desired axiomatic properties \cite{shapley1953value}. It is linear as a function of the model ($\alpha, \beta \in \mathbb{R}$): $\phi_{\alpha f +\beta g}(i) = \alpha \phi_f(i) + \beta \phi_g(i)$, and the ``centered'' learned model is additively separable with respect to the Shapley values: $(\bm{x})-\mathbb{E}_{\text{Pr}}[f(\bm{X})] = \sum_{i=1}^{d} \phi_f(i)$, which is known as the \emph{efficiency} property.

Like originally developed in game theory, the Shapley value is designed for one-dimensinal codomain of the function $f$. For explaining machine learning models which output multidimensional discrete probability distribution, like in multiclass classification, people have been explaining each output dimension one-by-one, applying a logit transformation to the probabilities, resulting in a one-vs-rest comparison. However, this ignores the relative information between each probability and the compositional nature of probability distributions. Indeed, the probabilistic output of a classifier lives on a multidimensional simplex. The latter is the sample space of data refered as \emph{compositional data} we briefly review in the next section.

@@

Compositional data carries relative information. Each element of a composition \emph{describes a part of some whole} \cite{pawlowskymodeling} like vectors of proportions, concentrations, and discrete probability distributions. A $N$-part composition is a vector of $N$ non-zero positive real numbers that sum to a constant $k$. Each element of the vector is a part of the \emph{whole} $k$. The sample space of compositional data is known as the simplex: $\mathcal{S}^N = \left\{ \bm{x} = [x_1, x_2,\dots x_{N}]^T \in \mathbb{R}^{*N}_{+} \big| \sum_{i=1}^{N} x_i = k \right\}$. In a composition, only the relative information between parts matters and John Aitchison introduced the use of log-ratios of components to handle this \cite{aitchison1982}. He defined several operations on the simplex which leads to what is called the \emph{Aitchison geometry of the simplex}.

\subsection{The Aitchison geometry of the simplex}
John Aitchison defined an internal operation called \emph{perturbation}, an external one called \emph{powering} and an inner product \cite{aitchison2001}:
\begin{itemize}
\item a \emph{perturbation}:
  $\bm{x}\oplus \bm{y} = \mathcal{C}\left([x_1y_1,\dots x_{N}y_{N}]\right)$ seen as an addition between two compositions $\bm{x},\bm{y}\in \mathcal{S}^N$,
\item a \emph{powering}:
  $\alpha \odot \bm{x} = \mathcal{C}\left([x_{1}^{\alpha},\dots x_{N}^{\alpha}]\right)$ seen as a multiplication by a scalar $\alpha \in \mathbb{R}$,
\item an inner product:\\
  $\displaystyle \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j}$.
\end{itemize}
$\mathcal{C}(\cdot)$ is the closure operator. Since only the relative information matter, scaling factors are irrelevant and a composition $\bm{x}$ is equivalent to $\lambda \bm{x} = [\lambda x_1,\lambda x_2,\dots\lambda x_N]$ for all $\lambda>0$. This equivalence is materialized by the closure operator defined for $k>0$ as: $\mathcal{C}\left(\bm{x} \right) = \left[ \frac{k x_1}{\lVert \bm{x} \rVert_1}, \frac{k x_2}{\lVert \bm{x} \rVert_1} ,\dots \frac{k x_N}{\lVert \bm{x} \rVert_1} \right]^T$, where $\bm{x} \in \mathbb{R}_+^{*N}$ and $\displaystyle \lVert \bm{x} \rVert_1 = \sum_{i=1}^N \lvert x_i \rvert$.%Therefore, any vector of positive real numbers can be projected onto the simplex using the closure.

@@

Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express its ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$. The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex. The choice of the basis will be discussed in Section \ref{sec:balances}.

@@

In Section \ref{sec:shapley} we have briefly presented the standard formulation of Shapley value designed for one-dimensional prediction. In this Section, we will see how the Aitchison geometry can be used to extend this concept to the multidimensional simplex for explaning probabilistic predictions.

Let $\bm{f}:\mathcal{X}\to\mathcal{S}^N$ be a learned model, like a $N$-classes probabilistic classifier for instance, which outputs a probabilistic prediction on the $(N-1)$-dimensional probability simplex $S^N$. To properly consider the relative information between the probabilities, The outputs of the model must be treated as compositional data using the operators and metric defined by the Aitchison geometry of the simplex. We therefore rewrite the contribution and the value function of Equations \ref{eq:contrib} and \ref{eq:valuefunction} as follow:
\begin{equation}
  \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) = \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S),
\end{equation}
where $\bm{a}\ominus\bm{b}$ is the perturbation $\bm{a} \oplus \left( (-1)\odot \bm{b}\right)$ which correspond to a substraction between compositions $\bm{a}$ and $\bm{b}$, and where:
\begin{equation}
  \label{eq:valuefunctionsimplex}
  \begin{aligned}
    \bm{v}_{\bm{f},\bm{x},\text{Pr}}: 2^{\mathcal{X}} &\to \mathcal{S}^N,\\
    \bm{X}_S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S].
  \end{aligned}
\end{equation}
The $\mathcal{A}$ in superscript highlight the fact that the expectation is done with respect to the Aitchison measure, rather than the Lebesgue measure, which can simply be computed as \cite{pawlowskymodeling}: $\mathbb{E}^{\mathcal{A}}[\bm{Y}] = \ilr^{-1}\left( \mathbb{E} \left[ \ilr\left( \bm{Y} \right) \right] \right)$,
where $\mathbb{E}^{\mathcal{A}}$ refers to the expectation with respect to the Aitchison measure while $\mathbb{E}$ refers to the expectation with respect to the Lebesgue measure.

@@

The Shapley compositions can be visualized in the Euclidean space isometric to the simplex thanks to the ILR transformation presented in Section \ref{sec:ilr}. This space has the advantage of being intuitive since it is the standard real $(N-1)$-dimensional vector space we are used too.

@@

In the three classes case, the space is $2$-dimensional. We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width. The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to. In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier. Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region. Being orthogonal to the virginica class-composition, this suggest that this features does not contribute on the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class. The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
\begin{figure}

@@

\vspace{0.5cm}
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
    \caption{Sum of the Shapley compositions in the ILR space from the base to the prediction.}
    \label{fig:3classesshapsum}
  \end{subfigure}

@@

In a four classes example, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset \cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3. Since they are 64 pixels as a set of features, which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for better clarity. We again use a SVM classifier with an rbf kernel. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class-compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
  \label{fig:4classesshapsum}
\end{figure}

@@

When more than three classes are involved, all the dimensions of the ILR space cannot be visualized at once, but $2$ or $3$-dimensional subspaces can still be visualized. In order to select the ILR components to visualize, one needs to understand what they refer to. In this section, we briefly discuss the interpretation of the ILR components in our context.% of the explanation of a classifier prediction with more than three classes.

@@

This can be illustrated by a sequential binary partition or bifurcation tree. Figure \ref{fig:trees} gives two examples. \ref{fig:bifurc1} shows the bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$. The first balance $\tilde{p}_1$ first compare the probabilities of class $1$ and $2$. Each next balance then recursively compares the probability for the next class with the probabilities for the previous ones independently of all the others.

In some applications, one may be interested in particular comparisons of groups of classes. For instance, like in an example presented in \cite{egozcue2005groups}, if one wants to compare political parties or groups, it may be pertinent to have a balance comparing left and right-wing groups. But sometimes, there are no obvious relevant comparisons to study. For instance, in the handwritten digit recognition problem, it may first seem natural to compare odd with even numbers or prime with non-prime. However, being basically a shape recognition problem and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent. Be that as it may, the choice of the basis must be left open, whether or not it is based on a relevant strategy. Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}. We comment, for conciseness, a single $2$-dimensional subspace. More visualisation can be found in Appendix \ref{app:...}. Let's have a look at the the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$) as in Figure \ref{fig:moreclasses35}. It is like saying we are only interested in comparing the output probability for class 0 and for class 6, and in comparing the probability for class 1 with the group of probabilities for classes 7 and 8. $\tilde{p}_3$ depends only on the probability for the digits 0 and 6 and $\tilde{p}_5$ depends only on the probabilities for the digits 1, 7 and 8. Therefore, the class-composition for the others digits have a zero projection within this subspace and are not drawn in Figure \ref{fig:moreclasses35}. The projection of the class-compositions for 0 and 6 are orthogonal to the ones for class 1, 7 and 8. Indeed, the groups of digits making the balance $\tilde{p}_3$ and the groups of classes making $\tilde{p}_5$ have no intersection contrary to the example of Figure \ref{fig:3classes} where $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor. In this case, none of the class-compositions are orthogonal. In Figure \ref{fig:moreclasses35}, since $\tilde{p}_5$ is comparing 1 with the group of digits 7 and 8, the projection of the class-compositions on this line for 1 goes in an opposite direction than the one for the class-compositions for 7 and 8. The two latters, are equal and are half as long as the former. In this way, $\tilde{p}_5$ compares the probability for 1 with the group of probabilities for 7 and 8 with the same weight.
\begin{figure}
  \begin{subfigure}{0.5\textwidth}
    \centering
    \input{bifurctree.tex}
    \caption{Bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} and used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}.}
    \label{fig:bifurc1}
  \end{subfigure}

  \vspace{0.5cm}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \input{bifurctree1.tex}
      \caption{Bifurcation tree used in our $10$-classes digit recognition task}
      \label{fig:bifurc2}
    \end{subfigure}
  \caption{Two examples of bifurcation tree.}
  \label{fig:trees}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{figures/moreclasses/ilrplot35.png}
  \caption{Sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$.}
  \label{fig:moreclasses35}
\end{figure}
\subsection{Angles, norms and projections}

>>>>

Let $f:\mathcal{X}\to\mathbb{R}$ be a learned model one want to \emph{locally} explain where $f(\bm{x})$ is the prediction on the instance $\bm{x}\in\mathcal{X}\subset\mathbb{R}^d$. Let $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data\footnote{Usually, this is unknown but expectations will be replaced by empirical averagings.}. Let $S\subseteq \mathcal{I}=\{1,2,\dots d\}$, where $d$ is the number of features that composes an instance $\bm{x}\in\mathcal{X}$, be a subset of indices. $\bm{x}_S$ refers to an instance $\bm{x}$ restricted to the features indicated by the indices in $S$.

When an instance $\bm{x}$ is observed, the expected value of the prediction is simply $\mathbb{E}[f(\bm{x}) \mid \bm{x}] = f(\bm{x})$. However, when only $\bm{x}_S$ is given with $\mathcal{S} \neq \mathcal{I}$, there is uncertainty about the other features and we therefore compute the expected prediction given $\bm{x}_S$: $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$. The contribution of the feature indexed by $i \notin S$ in the prediction $f(\bm{x})$ given the known features indexed by $S$ is given by:
\begin{equation}
  \label{eq:contrib}
  c_{f,\bm{x},\text{Pr}}(i,\bm{X}_S) = v_{f,\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) - v_{f,\bm{x},\text{Pr}}(\bm{X}_S),
\end{equation}
where $v$ is known as the value function:
\begin{equation}
  \label{eq:valuefunction}
  \begin{aligned}
    v_{f,\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathbb{R},\\
    S &\mapsto \mathbb{E}_\text{Pr}[f(\bm{x})\mid \bm{x}_S],
  \end{aligned}
\end{equation}
where $2^{\mathcal{I}}$ is the set of all subsets of $\mathcal{I}$. This measure the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$. The whole contribution of the $i$th feature is computing by averaging this quantity over all possible coalitions as follow:
\begin{equation}
  \phi_{f,\bm{x},\text{Pr}}(i) = \frac{1}{d!} \sum_{\pi}c_{f,\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}),
\end{equation}
where $\pi$ is a permutation of the set $\mathcal{I}$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$. For better clarity, the subscript $_{f,\bm{x},\text{Pr}}$ wil be dropt from the equations.

This quantity is known as the Shapley value for the $i$th feature. It comes from cooperative game theory and is known to be the only quantity respecting a set of desired axiomatic properties \cite{shapley1953value}. It is linear as a function of the model ($\alpha, \beta \in \mathbb{R}$): $\phi_{\alpha f +\beta g}(i) = \alpha \phi_f(i) + \beta \phi_g(i)$, and the ``centered'' learned model is additively separable with respect to the Shapley values:
\begin{equation}
  f(\bm{x})-\mathbb{E}_{\text{Pr}}[f(\bm{X})] = \sum_{i=1}^{d} \phi_f(i),
\end{equation}
which is known as the \emph{efficiency} property.

The Shapley value is designed for one-dimensinal codomain of the function $f$. For explaining machine learning models which output multidimensional discrete probability distribution, like in multiclass classification, people have been explaining each output dimension one-by-one, applying a logit transformation to the probabilities, resulting in a one-vs-rest comparison. However, this approach ignores the relative information between each probability and ignores the compositional nature of probability distributions. Indeed, the probabilistic output of a classifier lives on a multidimensional simplex. The latter is the sample space of data refered as \emph{compositional data} we briefly review in the next section.

@@

Compositional data carries relative information. Each element of a composition \emph{describes a part of some whole} \cite{pawlowskymodeling} like vectors of proportions, concentrations, and discrete probability distributions. A $N$-part composition is a vector of $N$ non-zero positive real numbers that sum to a constant $k$. Each element of the vector is a part of the \emph{whole} $k$. The sample space of compositional data is the simplex: $\mathcal{S}^N = \left\{ \bm{x} = [x_1, x_2,\dots x_{N}]^T \in \mathbb{R}^{*N}_{+} \big| \sum_{i=1}^{N} x_i = k \right\}$. In a composition, only the relative information between parts matters and John Aitchison introduced the use of log-ratios of components to handle this \cite{aitchison1982}. He defined several operations on the simplex which leads to what is called the \emph{Aitchison geometry of the simplex}.

\subsection{The Aitchison geometry of the simplex}
John Aitchison defined an internal operation called \emph{perturbation}, an external one called \emph{powering} and an inner product \cite{aitchison2001}:
\begin{itemize}
\item a \emph{perturbation}:
  $\bm{x}\oplus \bm{y} = \mathcal{C}\left([x_1y_1,\dots x_{N}y_{N}]\right)$ seen as an addition between two compositions $\bm{x},\bm{y}\in \mathcal{S}^N$,
\item a \emph{powering}:
  $\alpha \odot \bm{x} = \mathcal{C}\left([x_{1}^{\alpha},\dots x_{N}^{\alpha}]\right)$ seen as a multiplication by a scalar $\alpha \in \mathbb{R}$,
\item an inner product:\\
  $\displaystyle \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j}$,
\end{itemize}
where $\mathcal{C}(\cdot)$ is the closure operator. Since only the relative information matter, scaling factors are irrelevant and a composition $\bm{x}$ is equivalent to $\lambda \bm{x} = [\lambda x_1,\lambda x_2,\dots\lambda x_N]$ for all $\lambda>0$. This equivalence is materialized by the closure operator defined for $k>0$ as: $\mathcal{C}\left(\bm{x} \right) = \left[ \frac{k x_1}{\lVert \bm{x} \rVert_1}, \frac{k x_2}{\lVert \bm{x} \rVert_1} ,\dots \frac{k x_N}{\lVert \bm{x} \rVert_1} \right]^T$, where $\bm{x} \in \mathbb{R}_+^{*N}$ and $\displaystyle \lVert \bm{x} \rVert_1 = \sum_{i=1}^N \lvert x_i \rvert$.%Therefore, any vector of positive real numbers can be projected onto the simplex using the closure.

@@

Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express its ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$. The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where the set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex. The choice of the basis will be discussed in Section \ref{sec:balances}.

@@

In Section \ref{sec:shapley} we have briefly presented the standard use of Shapley value designed for explaining one-dimensional predictions. In this Section, we will see how the Aitchison geometry can be used to extend the concept of Shapley value to the simplex for explaning multidimensional probabilistic predictions.

Let $\bm{f}:\mathcal{X}\to\mathcal{S}^N$ be a learned model, like a $N$-classes probabilistic classifier for instance, which outputs a probabilistic prediction on the $(N-1)$-dimensional probability simplex $S^N$. To properly consider the relative information between the probabilities, the output of the model must be treated as compositional data using the operators and metric defined by the Aitchison geometry of the simplex. We therefore rewrite the contribution and the value function of Equations \ref{eq:contrib} and \ref{eq:valuefunction} as follow:
\begin{equation}
  \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) = \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S),
\end{equation}
where $\bm{a}\ominus\bm{b}$ is the perturbation $\bm{a} \oplus \left( (-1)\odot \bm{b}\right)$ which correspond to a substraction between compositions $\bm{a}$ and $\bm{b}$, and where:
\begin{equation}
  \label{eq:valuefunctionsimplex}
  \begin{aligned}
    \bm{v}_{\bm{f},\bm{x},\text{Pr}}: 2^{\mathcal{X}} &\to \mathcal{S}^N,\\
    \bm{X}_S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S].
  \end{aligned}
\end{equation}
The $\mathcal{A}$ in superscript highlight the fact that the expectation is done with respect to the Aitchison measure, rather than the Lebesgue measure, which can simply be computed as: $\mathbb{E}^{\mathcal{A}}[\bm{Y}] = \ilr^{-1}\left( \mathbb{E} \left[ \ilr\left( \bm{Y} \right) \right] \right)$,
where $\mathbb{E}^{\mathcal{A}}$ refers to the expectation with respect to the Aitchison measure while $\mathbb{E}$ refers to the expectation with respect to the Lebesgue measure \cite{pawlowskymodeling}.

@@

The Shapley compositions can be visualized in the Euclidean space isometric to the simplex thanks to the ILR transformation presented in Section \ref{sec:ilr}. This space has the advantage of being intuitive since it is a standard real $(N-1)$-dimensional vector space.

@@

In the three classes case, the space is $2$-dimensional. We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width. The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to. In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier. Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most on the prediction, and moves the base distribution to the versicolor maximum probability decision region (maximum probability decision region boundaries are the dashed gray lines). Being orthogonal to the virginica class-composition, this suggest that this features does not contribute on the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class. The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
\begin{figure}

@@

\vspace{0.5cm}
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
    \caption{Sum of the Shapley compositions in the ILR space from the base distribution to the prediction.}
    \label{fig:3classesshapsum}
  \end{subfigure}

@@

In a four classes example, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset \cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3. Since they are 64 pixels, considering each pixel as a feature would correspond to 64 Shapley composition. We therefore reduce the number of features to 6 using a principal component analysis for better clarity and conciseness. We again use a SVM classifier with a rbf kernel. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class-compositions in the ILR space are the same as the bonds between the carbon and a hydrogen: the angles are $\approx 109.5^{\circ}$. In this example, the tested instance is a 0\footnote{More examples can be obtained from the notebooks released on the git repository: link not shared during the double blind peer reviewing.}.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction. The gray transparent walls markout the maximum probability decision regions.}
  \label{fig:4classesshapsum}
\end{figure}

@@

When more than three classes are involved, all the dimensions of the ILR space cannot be visualized at once, but $2$ or $3$-dimensional subspaces can still be visualized. In order to select the ILR components to visualize, one needs to understand what they refer to. In this section, we briefly discuss the interpretation of the ILR components.% of the explanation of a classifier prediction with more than three classes.

@@

This can be illustrated by a sequential binary partition or bifurcation tree. Figure \ref{fig:trees} gives two examples: \ref{fig:bifurc1} shows the bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$. The first balance $\tilde{p}_1$ first compare the probabilities of class $1$ and $2$. Each next balance then recursively compares the probability for the next class with the probabilities for the previous ones independently of all the others.

In some applications, one may be interested in particular comparisons of groups of classes. For instance, like in an example presented in \cite{egozcue2005groups}, if one wants to compare political parties or groups, it may be pertinent to have a balance comparing left and right-wing groups. However, there are sometimes no obvious relevant comparisons to study. For instance, in the handwritten digit recognition problem, it may first seem natural to compare odd with even numbers or prime with non-prime but being basically a shape recognition problem, and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent. Be that as it may, the choice of the basis must be left open, whether or not it is based on a relevant strategy. Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the Mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}. We comment, for conciseness, a single $2$-dimensional subspace\footnote{More visualisations can be obtaned from the shared python notebooks.}. Let's have a look, in Figure \ref{fig:moreclasses35}, at the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$). It is like saying we are only interested in comparing the output probability for class 0 and for class 6, and in comparing the probability for class 1 with the group of probabilities for classes 7 and 8. $\tilde{p}_3$ depends only on the probability for the digits 0 and 6 and $\tilde{p}_5$ depends only on the probabilities for the digits 1, 7 and 8. Therefore, the class-composition for the others digits have a zero projection within this subspace and are not drawn in Figure \ref{fig:moreclasses35}. The projection of the class-compositions for 0 and 6 are orthogonal to the ones for class 1, 7 and 8. Indeed, the set of classes making the balance $\tilde{p}_3$ and the set of classes making $\tilde{p}_5$ have no intersection contrary to the example of Figure \ref{fig:3classes} where $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor. In this case, none of the class-compositions are orthogonal. In Figure \ref{fig:moreclasses35}, since $\tilde{p}_5$ is comparing 1 with the group of digits 7 and 8, the projection of the class-compositions on this line for 1 goes in an opposite direction than the one for the class-compositions for 7 and 8. The two latters, are equal and are half as long as the former. In this way, $\tilde{p}_5$ compares the probability for 1 with the group of probabilities for 7 and 8 with the same weight.

Within this space, Shapley compositions can be explored like in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum} keeping in mind that this is a subset of the full ILR space. 
\begin{figure}
  \begin{subfigure}{0.5\textwidth}
    \centering
    \input{bifurctree.tex}
    \caption{Bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} and used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}.}
    \label{fig:bifurc1}
  \end{subfigure}

  \vspace{0.5cm}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \input{bifurctree1.tex}
      \caption{Bifurcation tree used in our $10$-classes digit recognition task}
      \label{fig:bifurc2}
    \end{subfigure}
  \caption{Two examples of bifurcation tree.}
  \label{fig:trees}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{figures/moreclasses/ilrplot35.png}
  \caption{Sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$ for a test instance from the class 2. $\tilde{p}_3$ compares the probability for the class 0 with the one for the class 6 and $\tilde{p}_5$ compares the probability for the class 1 with the group of probabilities for the class 7 and 8. }
  \label{fig:moreclasses35}
\end{figure}
\subsection{Angles, norms and projections}

Some may find the analysis of the features contributions in cases with more than four classes tricky. Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:balances}, choosing which subspaces to visualize require a careful understanding of the balances and a careful building of the bifurcation tree. However, the Shapley explaination can be summarized by sets of angles, norms and projections. The norm of a Shapley composition gives the strength of the feature's contribution in the prediction. It gives the overall contribution of the feature on the prediction, regardless of its direction. The angle between two Shapley compositions can informs about their orthogonality. If two Shapley composition are orthogonal%\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}
, this suggest the features are nonredundant. A negative angle would suggest that the features have an opposite influence on the prediction. The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing. Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, angles between them and their projection on the class-compositions.
